---
title: "lab_12"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   tidytext and our usual libraries

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
library(tidyverse)
library(tidytext)
library(janitor)
library(lubridate)
library(rvest)
```

## Questions

**Q1.** You've been assigned to report a story about the leading reasons that Maryland attorneys get sanctioned by the state for misconduct. The state [publishes lists of sanctions](https://www.courts.state.md.us/attygrievance/sanctions) that contain a short text description about the situation. Load the CSV file in the data folder containing records from fiscal year 2011 onwards. Make a list of unique words from the text column, then following the example in the pre_lab, remove common "stop words" from that list and create a list of the top 10 words containing the percentage of occurrences each word represents. What's the leading word in that answer and, broadly, what do you think the top 10 words describe?

**A1.** The leading word in the dataset is “failing,” which appears far more often than any other term. Broadly, the top 10 words point to a consistent pattern in attorney misconduct: problems with handling client funds (“trust,” “account,” “funds”), breaches of professional duty (“failing,” “failed,” “conduct,” “engaging”), and the disciplinary actions that follow (“consent,” “reprimand,” “representation”). Taken together, these terms show that many sanctions stem from attorneys not safeguarding money, not fulfilling promised services, or engaging in dishonest behavior.

```{r}

md_attorney_sanctions <- read_csv("data/md_attorney_sanctions.csv") |>
  clean_names()

unique_words <- md_attorney_sanctions |>
  select(text) |>
  unnest_tokens(word, text)

stop_words <- stop_words |>
  add_row(word = "respondent") |>
  add_row(word = "maryland") |>
  add_row(word = "client") |>
  add_row(word = "clients") |>
  add_row(word = "bar") |>
  add_row(word = "counsel") |>
  add_row(word = "attorney") |>
  add_row(word = "attorneys") |>
  add_row(word = "court") |>
  add_row(word = "commission")

top_words <- unique_words |>
  anti_join(stop_words, by = "word") |>
  group_by(word) |>
  tally(sort = TRUE) |>
  mutate(percent = (n / sum(n)) * 100) |>
  top_n(10)

top_words


```

**Q2.** Let's move beyond single words to phrases. Make a list of the top 10 three-word phrases, called trigrams, based on the example from the pre_lab (you'll need to modify the example code to do this). What's the top trigram and how often does it appear? What does that phrase mean in legal terms?

**A2.** The top trigrams all point to two major problems. The first three “conduct involving dishonesty,” “dishonesty fraud deceit,” and “involving dishonesty fraud” show that many sanctions are tied to lying, fraud, or other misleading behavior. The next two “trust account failing” and “collecting unreasonable fees” highlight issues with mishandling client money or charging improper fees. Together, the top ten trigrams make it clear that dishonesty and financial misconduct are the most common reasons attorneys get disciplined.

```{r}
trigrams <- md_attorney_sanctions |>
  select(text) |>
  unnest_tokens(trigram, text, token = "ngrams", n = 3) |>
  separate(trigram, c("word1", "word2", "word3"), sep = " ") |>
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |>
  filter(!word3 %in% stop_words$word) |>
  mutate(trigram = paste(word1, word2, word3, sep = " ")) |>
  group_by(trigram) |>
  tally(sort = TRUE) |>
  mutate(percent = (n / sum(n)) * 100) |>
  top_n(10)

trigrams

```

**Q3.** Let's drop back down to more traditional text analysis - take the top trigram from Q2 and write code to see how many times it occurs in the text column in each fiscal year. What do you think the answer produced by your code suggests? What else could you do to try and clarify the most important reasons attorneys get sanctioned?

**A3.** The top trigram appears only a few times in the earlier years but rises sharply starting around 2019 and peaks in 2021. This suggests that cases involving dishonesty became more common or more heavily emphasized in recent disciplinary actions.

```{r}

top_trigram <- trigrams$trigram[1]

trigram_by_year <- md_attorney_sanctions |>
  select(fiscal_year, text) |>
  unnest_tokens(trigram, text, token = "ngrams", n = 3) |>
  filter(trigram == top_trigram) |>
  group_by(fiscal_year) |>
  tally(name = "n_trigram") |>
  arrange(fiscal_year)

trigram_by_year
```
