---
title: "Reverse Engineering Project"
author: "Mehedi Marof_Group 4"
date: "10/22/2025"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this notebook, we are reverse engineering the story, [Thousands of kids are missing from school. Where did they go?](https://projects.apnews.com/features/2023/missing-children/index.html)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
# load your libraries
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data.

```{r}
# Load required data + clean column names
# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")
# Clean required data and prepare for analysis if needed. 

library(readr)
population_by_age <- read_csv("data/population-by-age.csv") |> clean_names()
private_enrollment_by_grade <- read_csv("data/private-enrollment-by-grade.csv") |> clean_names()
public_enrollment_by_grade <- read_csv("data/public-enrollment-by-grade.csv") |> clean_names()


```

## Sentences to Engineer

In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

-   **Sentence text**: “In the 21 states where data was available, enrollment fell by about 700,000 students between the 2019-2020 and 2021-2022 school years.”
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

enrollment_by_year <- private_enrollment_by_grade |>
  filter(sch_year == c("2019-2020", "2021-2022")) |>
  group_by(sch_year) |>
  summarise(total_enrollment = sum(enrollment, na.rm = TRUE)) |>
  arrange(sch_year)

enrollment_by_year

# calculate change in enrollment
enrollment_change <- enrollment_by_year |>
  summarise(change = total_enrollment[sch_year == "2021-2022"] -
                    total_enrollment[sch_year == "2019-2020"])

enrollment_change


# Display results of code below this codeblock

```

### Sentence 2

-   **Sentence text**: Across the states in the analysis, private-school enrollment grew by about 103,000 students.
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}

ap_states <- c("CA","CO","CT","DC","FL","GA","HI","IL","IN","IA",
               "KS","LA","MD","MA","MN","MO","NC","NE","NH","NJ","NY")

enrollment_by_year <- private_enrollment_by_grade |>
  filter(sch_year %in% c("2019-2020", "2021-2022"),
         state %in% ap_states) |>
  summarise(total_enrollment = sum(enrollment, na.rm = TRUE), .by = sch_year) |>
  arrange(sch_year)

enrollment_change <- enrollment_by_year |>
  summarise(change = total_enrollment[sch_year == "2021-2022"] -
                    total_enrollment[sch_year == "2019-2020"])

```

### Sentence 3

-   **Sentence text**: [Paste in sentence to engineer here]
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 4

-   **Sentence text**: [Paste in sentence to engineer here]
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

-   **Sentence text**: [Paste in sentence to engineer here]
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-
