---
title: "lab_04_ai"
author: "Daniel Trielli"
date: "2024-09-26"
output: html_document
---

## Setup

Let's load the necessary libraries and, using your API key, setup your credentials:

```{r}
library(axolotr)
create_credentials(GROQ_API_KEY = "")
```

See that "Please restart your R session for changes to take effect."? Go ahead and do that; you'll need to rerun the `library()` function for axolotr, and let's load tidyverse while we're at it.\

```{r}
library(axolotr)
library(tidyverse)
```

## Testing

Let's make sure that worked. We'll be using the [Llama 3.1 model released by Meta](https://ai.meta.com/blog/meta-llama-3-1/).

```{r}
groq_response <- axolotr::ask(
  prompt = "Give me a 100-word pitch for a new Lilo and Stitch sequel",
  model = "llama-3.1-8b-instant"
)

groq_response
```

Did that work?

## Q1. Turning unstructured information into data: let's take [this article](https://www.nytimes.com/2024/09/27/arts/maggie-smith-dead.html) about the death of Dame Maggie Smith and try to extract information from it. Your goal is to have Groq create a table with the films and TV shows mentioned in this news article, and extract the following information about them: title, year, role, director, co-stars, and awards. From the table that it created, answer this: is the information correct? Did all information come from the article text?

```{r}
#Loading article text
text = "Maggie Smith, one of the finest British stage and screen actors of her generation, whose award-winning roles ranged from a freethinking Scottish schoolteacher in /“The Prime of Miss Jean Brodie/” to the acid-tongued dowager countess on /“Downton Abbey,/” died on Friday in London. She was 89.

Her death, in a hospital, was announced by her family in a statement issued by a publicist. The statement gave no cause of death.

American moviegoers barely knew Ms. Smith (now Dame Maggie to her countrymen) when she starred in /“The Prime of Miss Jean Brodie/” (1969), about a 1930s girls’-school teacher who dared to have progressive social views — and a love life. Vincent Canby’s review in The New York Times described her performance as /“a staggering amalgam of counterpointed moods, switches in voice levels and obliquely stated emotions, all of which are precisely right./” It brought her the Academy Award for best actress.

She won a second Oscar, for best supporting actress, for /“California Suite/” (1978), based on Neil Simon’s stage comedy. Her character, a British actress attending the Oscars with her bisexual husband (Michael Caine), has a disappointing evening at the ceremony and a bittersweet night in bed.

In real life, prizes had begun coming Ms. Smith’s way in the 1950s, when at 20 she won her first Evening Standard Award. By the turn of the millennium, she had the two Oscars, two Tonys, two Golden Globes, half a dozen Baftas (British Academy of Film and Television Awards) and scores of nominations. Yet she could go almost anywhere unrecognized.

Until /“Downton Abbey./”

That series followed the Earl of Grantham (Hugh Bonneville), his mostly aristocratic family and his troubled household staff at their grand Jacobean mansion as the world around them, between 1912 and 1925, refused to stand still.

After its 2010 British premiere and its 2011 American debut, the show ran six seasons. Its breakout star, from the beginning, was Ms. Smith, playing Lord Grantham’s elderly and still stubbornly Victorian widowed mother, Violet Crawley, the dowager countess. She disapproved of electric lights, was unfamiliar with the word /“weekend/” and never met a person or situation she couldn’t ridicule with withering imperiousness. When her daughter-in-law considered sending a younger relative for a stay in New York, Lady Violet objected: /“Oh, I don’t think things are quite that desperate./”

Suddenly, in her mid-70s, Ms. Smith was a megastar.

/“It’s ridiculous. I’d led a perfectly normal life until ‘Downton Abbey,’ /” she told Mark Lawson at the B.F.I. and Radio Times Festival in 2017, adding later: /“Nobody knew who the hell I was./”

The closest Ms. Smith had come to such visibility was with the Harry Potter movies. She was Minerva McGonagall, the Hogwarts School’s stern but fearless transformation teacher, in seven of the eight films, from /“Harry Potter: The Sorceror’s Stone/” (2001) to /“Harry Potter: The Deathly Hallows Part 2/” (2011).

McGonagall, wearing high-necked Victorian-style gowns, a distinctive Scottish brooch, and upswept hair beneath a tall, black witch’s hat, was a striking onscreen presence. Yet Ms. Smith did not find herself constantly pursued in public, except by children.

/“A lot of very small people kind of used to say hello to me, and that was nice,/” she recalled on /“The Graham Norton Show/” in 2015. One boy carefully asked her, /“Were you really a cat?/”)"
```

```{r}
#Add code below to talk to Groq and display its response


q1_prompt <- paste(
  "You are an expert data-extraction assistant for investigative journalists.",
  "Your task: extract structured, factual information ONLY from the obituary text below.",
  "Identify every *film* or *scripted TV series* in which Maggie Smith had an acting role that is explicitly mentioned.",
  "",
  "Return the result as a **pure CSV string**, with the following exact headers:",
  "title,year,role,director,co_stars,awards,source_quote",
  "",
  "Output rules:",
  "- Include one row per unique work (combine sequels/series into a single row if the text refers to them collectively).",
  "- The 'year' field should reflect what the text states (single year, span, or premiere notes).",
  "- 'role' = the character or description used in the text (quote or paraphrase from text).",
  "- 'director' = leave blank unless explicitly named in the article.",
  "- 'co_stars' = only people explicitly mentioned as acting with her; separate multiple with semicolons.",
  "- 'awards' = only awards explicitly attributed to that work.",
  "- 'source_quote' = a short (<=20 words) snippet from the obituary proving the info.",
  "- Leave any field blank if the article doesn’t provide that information.",
  "- Exclude interviews or talk-show appearances (e.g., The Graham Norton Show).",
  "- Produce nothing except the CSV — no markdown, no commentary, no code fences.",
  "",
  "TEXT:\n", text
)

q1_csv <- axolotr::ask(
  prompt = q1_prompt,
  model = "llama-3.1-8b-instant"
)


library(readr)
maggie_df <- read_csv(q1_csv)


```

**Answer:** Groq correctly identified four works: The Prime of Miss Jean Brodie, California Suite, Harry Potter and the Sorcerer’s Stone, and Downton Abbey. The years and award information are accurate, and no extra films were added. However, there are some small issues. The Downton Abbey “director” field mistakenly lists “the dowager countess,” which is part of her role, not a director’s name. The Harry Potter entry should cover the full series from 2001 to 2011 instead of only the first film. The description “Michael Caine’s wife” is somewhat interpretive but still reflects what the article said. Overall, the table is mostly accurate and based on information actually in the article, with only minor errors.

------------------------------------------------------------------------

## Q2. Helping with code explanation: Your data journalism instructor has given you a block of code but you can't make sense of it. Ask Groq if it can explain what the code does with this UMD course. Is the answer correct?

```{r}
# Loading R code example
r_code_example = 'umd_courses |>
  filter(str_detect(title, "Climate") & seats > 0) |>
  group_by(department) |>
  summarize(classes = n()) |>
  arrange(desc(classes))'
```

```{r}
#Add code below to talk to Groq and display its response
q2_prompt <- paste(
  "You are an R and tidyverse teaching assistant for a data journalism course at the University of Maryland.",
  "Explain clearly what this code does, step by step, assuming the dataset is named umd_courses.",
  "Describe the purpose of each function and what the final output looks like.",
  "Keep the explanation factual, short, and in plain English so a student can understand it.",
  "CODE:\n", r_code_example
)

q2_response <- axolotr::ask(
  prompt = q2_prompt,
  model = "llama-3.1-8b-instant"
)

q2_response
```

**Answer:** Groq explained the code correctly. It said the code filters courses that have “Climate” in the title and seats greater than zero, then groups them by department, counts how many classes each department has, and sorts the result from most to least. The explanation was clear and easy to follow. The only small thing missing is that `str_detect()` comes from the stringr package. Overall, the answer was accurate.

------------------------------------------------------------------------

## Q3. Helping with code debugging: paste the code block of an answer you had for a previous lab down here and ask Grok if that code is correct, based on the question in the lab. What do you think about its response?

```{r}
#Loading lab question
lab_question = 'Using case_when(), create a column in the Maryland expenses data called spending_location indicating whether each record indicated money spent in Maryland or outside Maryland, based on the address column of the \'maryland_expenses\' dataset. For records that are in Maryland, make the new column\'s value \"In-state\" and for the others, make it \"Out of state\". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).'


#Paste the code block here, between the quotes. If your code has quotes or single quotes, you have to add a \ before each one so R doesn't break.

your_r_code_lab <- "
maryland_expenses <- maryland_expenses |>
  mutate(
    spending_location = case_when(
      str_detect(address, \"Maryland\") ~ \"In-state\",
      str_detect(address, \"MD\") ~ \"In-state\",
      .default = \"Out of state\"
    )
  )
  spending_by_location <- maryland_expenses |>
  group_by(spending_location) |>
  summarise(total_spent = sum((amount), na.rm = TRUE))
"
```

```{r}
#Add code below to talk to Groq and display its response
q3_prompt <- paste(
  "You are a strict R/tidyverse reviewer for a data journalism class.",
  "First, restate the requirement in one sentence.",
  "Then give a verdict: Meets / Does not meet, with a one-line reason.",
  "List up to 4 short bullets on issues/fixes (edge cases like NA addresses, case-insensitive matching, word boundaries).",
  "Finally, provide ONE corrected, runnable R code block using the |> pipe that:",
  "- creates spending_location via case_when() using only the address column,",
  "- treats 'MD'/'Maryland' case-insensitively with word boundaries,",
  "- handles NA addresses explicitly,",
  "- summarizes total amount by spending_location with na.rm = TRUE, and",
  "- returns a 2-column tibble (spending_location, total_amount).",
  "Keep the response concise.\n\n",
  "LAB REQUIREMENT:\n", lab_question, "\n\n",
  "STUDENT CODE:\n", your_r_code_lab
)

q3_response <- axolotr::ask(
  prompt = q3_prompt,
  model = "llama-3.1-8b-instant"
)

q3_response 

"



```

**Answer:** Groq’s response was clear and accurate. It correctly restated the goal of the assignment and gave a fair verdict. It pointed out that my code worked but didn’t handle NA addresses and didn’t check “MD” or “Maryland” in a case-insensitive way. The suggested fixes made sense: add `!is.na(address)`, use `tolower()` or regex for case insensitivity, and ensure the column name is consistent. The corrected code it provided would work as intended and better follows the assignment instructions.

------------------------------------------------------------------------

## Q4. Brainstorming about strategies for data analysis and visualization: ask Groq to give you ideas about potential news stories from data analyses using [this data set](https://data.montgomerycountymd.gov/Public-Safety/Crash-Reporting-Incidents-Data/bhju-22kf/about_data). You're going to have to describe the dataset so that Groq can make some suggestions. What do you think of its response?

```{r}
#Add code below to talk to Groq and display its response
q4_prompt <- paste(
  "You are a data editor at a newsroom helping a journalism student brainstorm story ideas.",
  "The dataset is the Montgomery County, MD 'Crash Reporting – Incidents Data' available at",
  "https://data.montgomerycountymd.gov/Public-Safety/Crash-Reporting-Incidents-Data/bhju-22kf/about_data.",
  "It includes fields such as date/time of crash, latitude/longitude, street names, injury severity,",
  "pedestrian/bicycle involvement, weather, lighting, road conditions, and contributing factors.",
  "",
  "Based on this dataset, generate 8–10 concrete and realistic story ideas that could be developed using data analysis and visualization.",
  "For each idea, include:",
  "1. The question or angle for the story.",
  "2. The relevant fields to use from the dataset.",
  "3. The type of analysis or comparison to run (e.g., grouping, mapping, regression, before/after).",
  "4. The chart, map, or visualization that would help tell the story.",
  "5. One caveat or limitation to consider (e.g., underreporting, geocoding errors, exposure).",
  "",
  "Write the ideas as concise bullet points suitable for a newsroom pitch list. Keep the tone factual and journalistic."
)

q4_response <- axolotr::ask(
  prompt = q4_prompt,
  model = "llama-3.1-8b-instant"
)

q4_response
```

**Answer:** Groq suggested ten story ideas: pedestrian safety hotspots, weather-related crashes, bike lane effectiveness, school zone safety, time-of-day crash patterns, weekend vs. weekday trends, alcohol-related crashes, lighting conditions and crash severity, high-risk intersections, and seasonal crash patterns.

The most interesting ones are:

**Weekend vs. Weekday and Alcohol-Related Crashes** – This story could see = how crash patterns differ between weekdays and weekends, especially when alcohol is involved. It could show if weekends see more severe or late-night crashes linked to drinking. Mapping weekend alcohol-related crashes and comparing them with weekday ones could reveal nightlife or high-risk areas. A simple bar chart or heat map would make this pattern clear and connect driving behavior to safety concerns.

**Lighting, Weather, and Seasonal Crash Patterns** – This story could look at how environmental factors like lighting, weather, and season affect crash severity. It could show whether dark, rainy, or winter conditions lead to more serious crashes. Comparing crash counts under different lighting and weather conditions across months could highlight clear risk patterns. Visuals like heat maps or seasonal trend lines would make the findings easy to understand and valuable for reporting on road safety and infrastructure improvements.

------------------------------------------------------------------------

## Q5. Using AI to write news articles: ask Groq to write a 500-word news article about a car crash at the intersection between Knox Road and Regents Drive. What issues or problems do you see with the output?

```{r}
#Add code below to talk to Groq and display its response

```

**Answer:**
