---
title: "lab_10"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries, plus some of the map-specific ones from the pre_lab. You'll also be fetching Census data in Q3. Read the questions carefully!

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Load libraries
library(tidyverse)
library(sf)
library(janitor)
library(tigris)
library(tidycensus)
#library(maptools)

```

Let's explore Maryland elections, power outages and electric vehicle data and make some maps!

## Questions

**Q1.** Q1. In 2022, Maryland elected the state's first two African-American statewide office holders: Wes Moore as governor and Anthony Brown as attorney general. You want to compare the two's electoral performances. Using the `md_statewide_county.csv` file in the data folder, calculate the total number of votes for both the governor's race and the attorney general's race, then percentages for both Moore and Brown *based on the total of their respective contests*. Then calculate the difference between their percentages.

Get the geometry for Maryland's counties using tigris and, joining that dataframe to your results dataframe, make a county-level map of the differences between Moore and Brown's percentages. What is the story here, and what is the county with the most interesting results?

**A1.** The map shows that Wes Moore generally performed slightly better than Anthony Brown across most Maryland counties. Moore had stronger support in some suburban and rural areas, while Brown’s results stayed more similar in heavily Democratic counties. The most interesting example is Anne Arundel County, where Moore’s margin was noticeably larger, suggesting some voters supported Moore for governor but didn’t support the Democratic candidate for attorney general at the same level.

```{r}

# Load election data
md_statewide_county <- read_csv("data/md_statewide_county.csv") |>
  clean_names()

# Calculate total Governor votes
gov_totals <- md_statewide_county |>
  mutate(total_gov = cox + moore + lashar + wallace + harding + sefcik + gov_writein,
         pct_moore = moore / total_gov * 100) |>
  select(geoid, county, pct_moore)

# Calculate total Attorney General votes
ag_totals <- md_statewide_county |>
  mutate(total_ag = peroutka + brown + ag_writein,
         pct_brown = brown / total_ag * 100) |>
  select(geoid, county, pct_brown)

# Combine and compute difference: Moore - Brown
moore_brown_diff <- gov_totals |>
  inner_join(ag_totals, by = c("geoid", "county")) |>
  mutate(diff_pct = pct_moore - pct_brown)

# Get Maryland counties geometry
md_counties <- counties(state = "MD", cb = TRUE) |>
  clean_names() |>
  rename(geoid = geoid, county = name)

# Converting geoid to character before joining
md_counties <- md_counties |>
  mutate(geoid = as.character(geoid))

moore_brown_diff <- moore_brown_diff |>
  mutate(geoid = as.character(geoid))

md_moore_brown_map <- md_counties |>
  inner_join(moore_brown_diff, by = "geoid")

# Map
ggplot(md_moore_brown_map) +
  geom_sf(aes(fill = diff_pct), color = "white", size = 0.3) +
  scale_fill_viridis_c(option = "magma", direction = -1,
                       name = "Moore – Brown (%)") +
  labs(
    title = "Difference in Vote Share: Wes Moore vs. Anthony Brown (2022)",
    subtitle = "Percent difference within each county (Moore % minus Brown %)",
    caption = "Source: Maryland 2022 Election Results"
  ) +
  theme_minimal()

```

**Q2.** Make a county-level map to explore the number of customers impacted by power outages in Maryland from last March. Load the `county_outages_march2023.csv` file in the data folder and calculate a rate of customers with outages per 10,000 total customers for each county. Then join that dataframe to your Maryland counties dataframe you made in Q1 and create a county-level map of the results; you may need to tweak the fill scale to make the distinctions clearer. What's the potential story here? Given access to other months' outage data, what would your next steps be in reporting a story?

**A2.** Talbot County has a much higher rate of outages than the rest of Maryland. Most other counties had very few outages. With more months of data, we could check if this keeps happening.

```{r}
# Load outages data
county_outages <- read_csv("data/county_outages_march2023.csv") |>
  clean_names() |>
  mutate(
    outage_rate_10k = outage_customers / total_customers * 10000
  )

# Join with Maryland counties geometry using county name
md_outages_map <- md_counties |>
  left_join(county_outages, by = c("county" = "county"))

# Plot the map
ggplot(md_outages_map) +
  geom_sf(aes(fill = outage_rate_10k), color = "white", size = 0.3) +
  scale_fill_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Outages per\n10,000 customers"
  ) +
  labs(
    title = "Power Outage Rates in Maryland (March 2023)",
    subtitle = "Customers with outages per 10,000 total customers",
    caption = "Source: Maryland outage data"
  ) +
  theme_minimal()



```

**Q3.** You've been assigned a story about the use of electric vehicles in Maryland and you want to make a map of the prevalence of electric vehicles in each county. Load the `ev_2020_2023.csv` file, which has the number of electric and hybrid vehicles per county from July 2020 and March 2023 and calculate a percentage change for each county (remember, that's `new-old/old`). Using the Maryland counties dataframe you made for Q1, make a county map showing the percentage change. What counties stand out to you? What questions do you have about the changes you're seeing?

Next, use tidycensus to retrieve county population data for Maryland from 2021 (you did this in the pre_lab) and join it to your vehicles dataframe. HINT: you may not need to join it on the county name, and you should already have the geometry!

Then, calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate. You will need to choose the "per capita" value.

Finally, make a second county map showing the per capita rate. What counties stand out on this map? How is it different from the first one you made?

**A3.** Montgomery, Howard, and Frederick counties saw the biggest increase in EVs. Some small counties also grew fast only because they had very few EVs before. In the per-capita map, wealthier suburban counties stand out, showing where EVs are actually more common, not just growing quickly.

```{r}
# Load EV data + percent change
ev <- read_csv("data/ev_2020_2023.csv") |>
  clean_names() |>
  mutate(
    pct_change = (march_2023 - july_2020) / july_2020 * 100
  )

# Join EV data to Maryland map by county name
ev_geo <- md_counties |>
  left_join(ev, by = c("county" = "county"))

ggplot(ev_geo) +
  geom_sf(aes(fill = pct_change), color = "white", size = 0.3) +
  scale_fill_viridis_c(option = "magma", direction = -1,
                       name = "% Change 2020–2023") +
  labs(
    title = "Percent Change in EVs (2020–2023)",
    subtitle = "Maryland Counties",
    caption = "Source: Maryland EV registrations"
  ) +
  theme_minimal()



```

```{r}
# Get Maryland population (2021) from ACS
md_pop <- get_acs(
  geography = "county",
  state = "MD",
  year = 2021,
  variables = "B01003_001"
) |>
  clean_names() |>
  select(geoid, population = estimate)

ev_geo <- ev_geo |>
  left_join(md_pop, by = "geoid") |>
  mutate(
    march_2023 = as.numeric(march_2023),
    population = as.numeric(population),
    ev_per_1000 = march_2023 / population * 1000
  )

ggplot(ev_geo) +
  geom_sf(aes(fill = ev_per_1000), color = "white", size = 0.3) +
  scale_fill_viridis_c(option = "viridis", direction = -1,
                       name = "EVs per 1,000") +
  labs(
    title = "Per Capita EVs in Maryland",
    subtitle = "EVs per 1,000 residents (2023)",
    caption = "Sources: MD EV registrations; ACS 2021"
  ) +
  theme_minimal()



```
